Live Functional Programming with Typed Holes

### 2018-04-17 examples-1.tex
sec2!

\subsection{Live Programming with Incomplete Expressions}

The process of writing new program fragments includes many states in which the
program under construction is, by definition, incomplete.
%
It is natural for a programmer to ``jump back and forth'' between different
places in the program, iteratively and alternately developing the missing
pieces.
%
The primary benefit of our approach is that such programs can be evaluated in
order to provide useful feedback during this editing workflow.

\input{fig-grades-example}

\overviewExample{1a}{Computing Weighted Averages}
%
Consider a teacher who is in the midst of developing a \Hazel{} notebook
(depicted at the top of \autoref{fig:grades-example}) to compute final student
grades at the end of a term.
%
At the top of the program, the teacher defines a
record type, \li{student_data}, for each student's course data.
%
In our simplified example, this includes the student's name, of type
\li{string}, and five grades, each of type \li{float}.

The teacher begins writing a \li{weighted_average} function that will
compute a weighted average, of type \li{float}, for each \li{student_data}
record.
%
The body of \li{weighted_average} is not yet written---marked with the
\emph{empty hole} on line \rkc{XXX}---but the teacher jumps ahead to map the
function over the \li{grades} list on line \rkc{XXX}.
%
\Hazel{} evaluates this incomplete program, showing its
\emph{indeterminate} result in \emph{results panel} just below the
program in the center of the window (\autoref{fig:grades-example});
this value is a list whose length is the same as
\li{grades}, where each element is the (indeterminate) result of evaluating the
hole expression inside \li{weighted_average}.
%
\rkc{The editor shows the first three elements of a result list, by default; the
... can be clicked and expanded to see additional elements.}

Notice that the hole expression on line \rkc{XXX} is assigned a unique static
identifier (\ie{}~\li{a}) and that the indeterminate results it produces are
assigned corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2},
etc.).
%
In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
using placeholder expressions like \li{raise "Not yet implemented."}, would halt
the program as soon it is evaluated (in this case, inside the definition of
\li{List.map} when the function is applied to the first element of \li{grades}.
%
By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
least see that the resulting expression does indeed evaluate to a list with the
correct number of (indeterminate) values.

Now the teacher returns to \li{weighted_average} to work on the missing
expression, with the goal of computing a weighted sum of each of a student's
grades.
%
The teacher performs several \Hazel{} edit actions (each of which is a single
keystroke, a shortcut to selecting one of the edit tools from the left panel,
followed by a tool-specific argument terminated with the Enter key):
%
navigating the cursor to hole \li{a} and typing \li{+}, which creates a
sum expression \li{??_b + ??_c} with the cursor placed, by default, under
left operand (the hole labeled \li{b});
%
typing \li{*} to create a multiplication expression \li{??_d * ??_e + ??_c},
with the cursor place under hole \li{d};
%
typing \li{\#10} to enter a numeric scaling factor, which then moves
the cursor to the second operand, hole \li{e}; and
%
typing \li{vg.hw1} to use the \li{hw1} field of the function argument
as the right operand of the multiplication.
%
The resulting partial expression, \li{(10.0 *. g.hw1) +. ??_b}---shown
in the screenshot in the bottom half of \autoref{fig:grades-example}---marks the
first of five weighted summands in the eventual finished \li{weighted_average}
calculation.

Each \Hazel{} edit action transforms an incomplete but well-formed and
well-typed program into another such program, and, thus, \Hazel{} immediately
runs the program after each edit.
%
After the above edit sequence, \Hazel{} display the updated list of
indeterminate expressions \li{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
%
Right away, the teacher recognizes that the values are too large; they should be
at most \li{100.0}.
%
The teacher realizes that representing percentage points as \li{float}s requires
that the constant on line \rkc{XXX} ought to be \li{0.10} instead.
%
Because of this live feedback, the teacher corrects this error right away and
avoids making similar programming errors in the rest of the calculation.
%
The teacher continues to build the rest of the arithmetic expression until it is
complete (there are no longer any expression holes), and the result of
immediately running the finished program shows that each of values in the final
result list is in the range \li{0.0} to \li{100.0}.

\input{fig-grades-example-2}

\overviewExample{1b}{Assigning Letter Grades}
%
The teacher's next task is to map the weighted averages to the letter grades A
through F (we consider only ``whole'' letter grades, for simplicity).
%
The \li{grade_cutoffs} record type, shown at the top of
\autoref{fig:grades-example-b}, describes the minimum cutoff for each of
these six possible grades.
%
Initially, each value in the \li{cutoffs} record is a hole.

%
%% initially all holes, because it will be different year-to-year based on the %
%data, differences in course difficulty, and to satisfy fairness criteria.
%
Before starting to fill in the cutoff values, the teacher jumps ahead to write a
function \li{letter_grade} that will make the connection between \li{cutoffs}
and \li{weighted_averages}.
%
Because she intends to look at the data to help select the cutoff values, the
teacher sorts the \li{weighted_averages} (on line \rkc{XXX}) and then maps
them to \li{letter_grades} (on line \rkc{XXX}).

When \Hazel{} runs this program, the guard of the outermost conditional
(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
is.
%
Therefore, each of the indeterminate expressions in \li{letter_grades} is the
entire expression body, albeit with different bindings for \li{n}.
%
Displaying such ``large'' indeterminate expressions can quickly consume all
available screen space, overwhelming the user with too much information.
%
\autoref{fig:XXX} shows how \Hazel{} renders large indeterminate
expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
or hole) or list of small values) simply as \li{..} to save space.
%
Hovering over this abbreviation (\rkc{???}) displays the full
indeterminate expression---as well as the evaluation environment that surrounds
it---as a tooltip.
%
\autoref{sec:discussion} discusses this and other user interface concerns when
trying to display useful live feedback without overwhelming the user.
%
These usability factors are beyond the scope of our work, which is to define
semantic foundations on which such user interfaces can be built.

To start deciding \li{cutoffs}, the teacher clicks the \li{weighted_averages}
expression, and views the results panel to see the data sorted in descending
order.
%
The result shows a natural gap between \li{92.2} and \li{89.5}.
%
So, she chooses to use \li{92.0} as the cutoff for A, replacing hole \li{XXX} on
line \rkc{XXX} with that numeric value.
%
Resuming the computation from before, \Hazel{} resolves the conditional
expressions for the first \rkc{XXX} indeterminate expressions, because each of
those \li{n} values was greater than \li{cutoffs.a}.
%
The remaining \rkc{XXX} expressions also proceeded to evaluate the first guard,
and are now indeterminate at the guard for the second conditional.

Before assigning subsequent cutoff values, the teacher would like to get a sense
of whether this first choice is a good one.
%
She jumps ahead to to write a function (\li{compute_distribution} on lines
rkc{XXX}-\rkc{XXX}) that computes the distribution of letter grades based
on \verb+cutoffs+..
%
Running \li{compute_distribution} shows that the percentage of As is
\rkc{XXX\%}, which is smaller than what the teacher would like;
the remaining percentages are all indeterminate.
%
Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
and then another gap between \li{88.2} and {85.5}.
%
So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
%
Because this edit is not just filling a hole expression, \Hazel{} discards the
previous execution state and reevaluates the entire program.
%
Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
be applied to seek opportunities for reuse even when non-hole expressions are
modified.
%
Because the focus of work is on the novel implications of running programs with
holes, our calculus and implementation supports caching ``edit-and-resume'' only
for the novel situation in which evaluation proceeds around holes.
%
After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
matches the teacher's intention.

In this fashion, the teacher continues down the list of sorted averages,
determining appropriate values for each cutoff.
%
Whenever the teacher is only filling in the ``next'' cutoff, the computation
from before can simply be resumed.
%
Overall, throughout the workflow described in these two examples, the programmer
can continue to evaluate the program, and receive meaningful feedback, while
going back and forth between different pieces under development.

>>>>

% !TEX root = hazelnut-dynamics.tex

\subsection{Example 1: Evaluating Past Holes and Hole Closures}

Consider the perspective of a teacher in the midst of developing a \Hazel{} notebook
to compute final student grades at the end of a course.
The top of \autoref{fig:grades-example} depicts the cell containing the incomplete program that 
the teacher has written so far (we omit irrelevant parts of the UI). 
%

At the top of this program, the teacher defines a
record type, \lismall{Student}, for recording a student's course data---here,
the student's name, of type
\lismall{string}, and, for simplicity, three grades, each of type \lismall{float}. The variable \lismall{students} binds a list of such student records. For simplicity, we include only three example students.
At the bottom of the program, the teacher maps a function \li{weighted_average} over this student data (\li{map} is the standard map function over lists, not shown), intending to compute a final weighted average for each student.
However, the program is incomplete because the teacher has not yet completed the body of the \lismall{weighted_average} function. This pattern is well-established: programmers often consume a function before implementing it.

Thusfar in Fig.~\ref{fig:grades-example}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator. The cursor now sits at an \emph{empty hole}, as indicated by the vertical bar and the green background. Each hole has a unique name (generated automatically in \Hazel), here simply \li{1}. 
%In this case, there is only one hole in the program, but in general, there may be many such holes.

Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer.
In response, the programmer might insert a temporary universal placeholder, e.g. by raising an exception as suggested in Sec.~\ref{sec:intro} (note however that Elm does not support exceptions, by design). This would cause typechecking to succeed. 
However, when the programming environment attempts to evaluate the program, 
evaluation would proceed only as far as the first \li{map} iteration, 
which would call into \li{weighted_average} and then fail when attempting to evaluate the hole.%
\footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.

\Hazel{} does not take this ``exceptional'' interpretation of holes. Instead, evaluation continues past the hole, treating it as an opaque expression of the appropriate type. The result, shown at the bottom of Fig.~\ref{fig:grades-cell-mockup}, is a list of length $3$, confirming that \li{map} does indeed behave as expected in this regard despite the teacher having provided an incomplete argument. Furthermore, each element of the resulting list has been evaluated as far as possible, i.e. the arithmetic expression \li{30.0*hw} has been evaluated for each corresponding value of \li{hw}, as expected. Evaluation cannot proceed any further because holes appear as addends. We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a hole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in critical positions.

Given this indeterminate result, the teacher might take notice of the magnitude of the numbers being computed, e.g. \li{2640.0} and \li{2280.0}, and
decide immediately that a mistake has been made: the teacher wants to compute a 
weighted average between \li{0.0} and \li{100.0}, and so the correct
constant is \li{0.30} rather than \li{30.0}. (There are of course other possible fixes.)

Although these observations might 
save only a small amount of time in this case, 
% relative to the 
% situation where the teacher notices this mistake only once  
% the \li{weighted_average} function has been completed, 
it demonstrates
the broader motivations of live programming: continuous feedback about the dynamic behavior of the 
program can help confirm the mental model 
that the programmer has developed (in this case, regarding the behavior of \li{map}), or else help quickly dispel 
inconsistencies between the programmer's expectations and the 
actual behavior of the program (in this case, the arithmetic expression).

% The result itself is not the only live feedback that \Hazel reports to the programmer.
% Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
\Hazel also helps programmers reason about the dynamic behavior of bound variables in scope at a hole via the \emph{live context inspector}, normally displays it as a sidebar but shown disembodied in three states in Fig.~\ref{fig:grades-sidebar}. 
In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments.

New to our approach are the values associated with each binding.
These values come from the environments associated with each hole instance in the result. In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}. An environment is a partial mapping of each variable in scope to a value. We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical foundations detailed in Sec.~\ref{sec:calculus}). The programmer can select different closures by clicking on a hole instance (here, instance \li{1:1} was selected by default, as indicated by the purple outline), or cycle rapidly through all of the closures for the hole at the cursor by clicking the arrows, e.g. $\blacktriangleright$, at the bottom of the context inspector, or corresponding hotkeys.

\vspace{-4px}
\subsection{Example 2: Recursive Functions}\label{sec:qsort1}
\vspace{-2px}

\input{qsort-example}

The first example was purposely simple. Let us briefly consider a second more sophisticated example of these concepts: an incomplete implementation of quicksort as a recursive function, shown in Fig.~\ref{fig:qsort-cell-mockup}. Here, the programmer has selected a pivot, split the remainder of the list relative to this pivot and made the appropriate recursive calls. A hole appears in the return position of the function body as the programmer contemplates how to combine the results from the recursive calls.

The programmer has called \li{qsort} on an example list. However, 
the indeterminate result in this case is rather uninteresting: it simply confirms that for this example, evaluation went through the recursive case of \li{qsort}, which ends with hole \li{1}. The live context inspector, on the other hand, provides a wealth of feedback about the dynamic behavior of this incomplete program. For example, the programmer can confirm that
the lists \li{smaller} and \li{bigger} are appropriate named.

The results from the recursive calls, \li{r_smaller} and \li{r_bigger}, are again hole instances, \li{1:2} and \li{1:3}. The programmer can click on these hole instances to reveal the environments at the corresponding recursive calls. For example, clicking on \li{1:2} reveals that the pivot in that case is \li{7} (not shown). In exploring these paths rooted at the result, the programmer can develop intuitions about the recursive structure of the computation before the program is complete. 
(We leave to future work the pedagogical question of whether such early feedback might help students with recursion.)

\begin{comment}
Now the teacher returns to \lismall{weighted_average} to work on the missing
expression, with the goal of computing a weighted sum of each of a student's
grades.
%
The teacher performs several \Hazel{} edit actions (each of which is a single
keystroke, a shortcut to selecting one of the edit tools from the left panel,
followed by a tool-specific argument terminated with the Enter key):
%
navigating the cursor to hole \lismall{a} and typing \lismall{+}, which creates a
sum expression \lismall{??_b + ??_c} with the cursor placed, by default, under
left operand (the hole labeled \lismall{b});
%
typing \lismall{*} to create a multiplication expression \lismall{??_d * ??_e + ??_c},
with the cursor place under hole \lismall{d};
%
typing \lismall{\#10} to enter a numeric scaling factor, which then moves
the cursor to the second operand, hole \lismall{e}; and
%
typing \lismall{vg.hw1} to use the \lismall{hw1} field of the function argument
as the right operand of the multiplication.
%
The resulting partial expression, \lismall{(10.0 *. g.hw1) +. ??_b}---shown
in the screenshot in the bottom half of \autoref{fig:grades-example}---marks the
first of five weighted summands in the eventual finished \lismall{weighted_average}
calculation.
\end{comment}

\begin{comment}
Each \Hazel{} edit action transforms an incomplete but well-formed and
well-typed program into another such program, and, thus, \Hazel{} immediately
runs the program after each edit.
%
After the above edit sequence, \Hazel{} display the updated list of
indeterminate expressions \lismall{[760.0 + ??_b.1, 880.0 + ??_b.2, ...]}.
%
Right away, the teacher recognizes that the values are too large; they should be
at most \lismall{100.0}.
%
The teacher realizes that representing percentage points as \lismall{float}s requires
that the constant on line \rkc{XXX} ought to be \lismall{0.10} instead.
%
Because of this live feedback, the teacher corrects this error right away and
avoids making similar programming errors in the rest of the calculation.
%
The teacher continues to build the rest of the arithmetic expression until it is
complete (there are no longer any expression holes), and the result of
immediately running the finished program shows that each of values in the final
result list is in the range \lismall{0.0} to \lismall{100.0}.
\end{comment}

\begin{comment}
\input{fig-grades-example-2}

\overviewExample{1b}{Assigning Letter Grades}
%
The teacher's next task is to map the weighted averages to the letter grades A
through F (we consider only ``whole'' letter grades, for simplicity).
%
The \lismall{grade_cutoffs} record type, shown at the top of
\autoref{fig:grades-example-b}, describes the minimum cutoff for each of
these six possible grades.
%
Initially, each value in the \lismall{cutoffs} record is a hole.

%
%% initially all holes, because it will be different year-to-year based on the %
%data, differences in course difficulty, and to satisfy fairness criteria.
%
Before starting to fill in the cutoff values, the teacher jumps ahead to write a
function \lismall{letter_grade} that will make the connection between \lismall{cutoffs}
and \lismall{weighted_averages}.
%
Because she intends to look at the data to help select the cutoff values, the
teacher sorts the \lismall{weighted_averages} (on line \rkc{XXX}) and then maps
them to \lismall{letter_grades} (on line \rkc{XXX}).

When \Hazel{} runs this program, the guard of the outermost conditional
(\lismall{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \lismall{cutoffs.a}
is.
%
Therefore, each of the indeterminate expressions in \lismall{letter_grades} is the
entire expression body, albeit with different bindings for \lismall{n}.
%
Displaying such ``large'' indeterminate expressions can quickly consume all
available screen space, overwhelming the user with too much information.
%
\autoref{fig:XXX} shows how \Hazel{} renders large indeterminate
expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
or hole) or list of small values) simply as \lismall{..} to save space.
%
Hovering over this abbreviation (\rkc{???}) displays the full
indeterminate expression---as well as the evaluation environment that surrounds
it---as a tooltip.
%
\autoref{sec:discussion} discusses this and other user interface concerns when
trying to display useful live feedback without overwhelming the user.
%
These usability factors are beyond the scope of our work, which is to define
semantic foundations on which such user interfaces can be built.

To start deciding \lismall{cutoffs}, the teacher clicks the \lismall{weighted_averages}
expression, and views the results panel to see the data sorted in descending
order.
%
The result shows a natural gap between \lismall{92.2} and \lismall{89.5}.
%
So, she chooses to use \lismall{92.0} as the cutoff for A, replacing hole \lismall{XXX} on
line \rkc{XXX} with that numeric value.
%
Resuming the computation from before, \Hazel{} resolves the conditional
expressions for the first \rkc{XXX} indeterminate expressions, because each of
those \lismall{n} values was greater than \lismall{cutoffs.a}.
%
The remaining \rkc{XXX} expressions also proceeded to evaluate the first guard,
and are now indeterminate at the guard for the second conditional.

Before assigning subsequent cutoff values, the teacher would like to get a sense
of whether this first choice is a good one.
%
She jumps ahead to to write a function (\lismall{compute_distribution} on lines
rkc{XXX}-\rkc{XXX}) that computes the distribution of letter grades based
on \verb+cutoffs+..
%
Running \lismall{compute_distribution} shows that the percentage of As is
\rkc{XXX\%}, which is smaller than what the teacher would like;
the remaining percentages are all indeterminate.
%
Returning to the value of \lismall{sorted_data}, she sees a cluster around \lismall{89.0}
and then another gap between \lismall{88.2} and {85.5}.
%
So, the teacher adjusts \lismall{cutoffs.a} to be \lismall{88.0}.
%
Because this edit is not just filling a hole expression, \Hazel{} discards the
previous execution state and reevaluates the entire program.
%
Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
be applied to seek opportunities for reuse even when non-hole expressions are
modified.
%
Because the focus of work is on the novel implications of running programs with
holes, our calculus and implementation supports caching ``edit-and-resume'' only
for the novel situation in which evaluation proceeds around holes.
%
After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
matches the teacher's intention.

In this fashion, the teacher continues down the list of sorted averages,
determining appropriate values for each cutoff.
%
Whenever the teacher is only filling in the ``next'' cutoff, the computation
from before can simply be resumed.
%
Overall, throughout the workflow described in these two examples, the programmer
can continue to evaluate the program, and receive meaningful feedback, while
going back and forth between different pieces under development.
\end{comment}

### 2018-04-17 examples-2.tex
sec2!

%% \subsection{Live Programming with Type Errors}
\subsection{Live Programming with Static Type Errors}

In the previous section, we described example programs that were incomplete
because of \emph{missing} expressions.
%
Next, we describe example programs that are incomplete because of
\emph{type-inconsistent} expressions.
%
For example, consider the following two definitions.

\begin{lstlisting}
let bad_bool : bool = ?? 0 ??_bad_bool;

let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
\end{lstlisting}

\noindent
%
These two definitions are ill-typed under standard typing disciplines.
%
In contrast, \citet{popl-paper} present a bidirectional type system that assigns
types to both, by wrapping type-inconsistent expressions (\li{0} on line
\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
%
Non-empty holes prevent local type inconsistencies from polluting the rest of
the program surrounding it, which may or may not itself contain additional
inconsistencies.

Understanding and debugging static type errors is notoriously difficult,
particularly for novices.
%
A variety of approaches have been
proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
and explain type errors.
%
One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
that demonstrates a run-time failure, and then displays a compressed execution
trace to the user as a graph.
%
In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
non-empty holes) as far as possible, in particular, until they
would go wrong.
%
Thus, our operational semantics provides a similar benefit to the approach of
\citet{Seidel2016}.

\overviewExample{3}{Sum List}
%
Consider the following buggy program (observed during an undergraduate
functional programming course~\cite{Seidel2016}) that attempts to sum a list
integers.
%
The error is that the base case produces a list rather than an integer.

@@

\noindent
%
Because the list expression on line 2 does not have type \li{int} as required,
it is wrapped in a (non-empty) hole by the bidirectional type
checker~\cite{popl-paper}.
%
Rather than trying to debug the error based on the static error, the programmer
may wish to trying running the function anyway by calling, say, \li{sum_list(2)}.
%
\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
%
By observing that the hole expression is being added to the integer \li{3}, he
realizes that it needs to be an integer, specifically, \li{0}.
%
Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
successfully proceeded to evaluate despite the error elsewhere.

@@

@@

>>>>

% !TEX root = hazelnut-dynamics.tex

%% \subsection{Live Programming with Type Errors}

\subsection{Example 3: Live Programming with Static Type Errors}

% \begin{subfigure}[t]{\textwidth}
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth,interpolate=false,valign=t]{images/qsort-type-error.png}
\caption{Example 3: Ill-Typed Quicksort}
\label{fig:qsort-type-error}
\vspace{-8px}
\end{figure}
% \end{subfigure}

The previous examples were incomplete 
because of \emph{missing} expressions.
%
Now, we discuss programs that are incomplete, 
and therefore conventionally meaningless, because of
\emph{type inconsistencies}. 
%
Let us return to the quicksort example just described, 
but assume that the programmer has filled in the previous hole
as shown in Fig.~\ref{fig:qsort-type-error}. (In Sec.~\ref{sec:resumption}, we discuss
how the programming environment might avoid restarting evaluation after such edits, but for small examples like this, restarting works fine.)

The programmer appears to be on the right track conceptually
in recognizing that the pivot needs to appear between the 
smaller and bigger elements. 
However, the types do not quite work out: the \li{@} operator here
performs list concatenation, but the pivot is an integer. 
Most compilers and editors will report a static error message, 
of varying utility, to the programmer in this case (and \Hazel 
follows suit, not shown). 
Our argument is that this should not cause all feedback about 
the dynamic behavior of the program to ``flicker out'' or ``go stale''. After all,
the error is localized and there is perfectly good code elsewhere 
in the program (if not nearby, then perhaps far away).

Our approach, following the prior work of \citet{popl-paper}, 
is to semantically internalize the ``red outline'' around
type inconsistencies as a \emph{non-empty hole}.
From the outside, a non-empty hole behaves statically much like an empty hole.
Our approach takes the same approach for the dynamic semantics,
i.e. evaluation safely proceeds past a non-empty hole just as if it were an empty hole.
The semantics also associates an environment with each instance of a non-empty hole (not shown). 
The difference is that evaluation also proceeds inside the hole, so that 
feedback about an expression that might ``almost'' be correct is available. 
In this case, the result at the bottom of Fig.~\ref{fig:qsort-type-error}
reveals that the programmer is on the right track: the list elements 
(here, for a shorter example for the sake of exposition) are in the right order.
They simply have not been combined correctly.

% For example, consider the following two definitions.

% \begin{lstlisting}
% let bad_bool : bool = ?? 0 ??_bad_bool;

% let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
% \end{lstlisting}

% \noindent
% %
% These two definitions are ill-typed under standard typing disciplines.
% %
% In contrast, \citet{popl-paper} present a bidirectional type system that assigns
% types to both, by wrapping type-inconsistent expressions (\li{0} on line
% \rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
% %
% Non-empty holes prevent local type inconsistencies from polluting the rest of
% the program surrounding it, which may or may not itself contain additional
% inconsistencies.

\begin{comment}
\overviewExample{2}{Sum List}
%
Consider the following buggy program (observed during an undergraduate
functional programming course~\cite{Seidel2016}) that attempts to sum a list
integers.
%
The error is that the base case produces a list rather than an integer.

@@

\noindent
%
Because the list expression on line 2 does not have type \li{int} as required,
it is wrapped in a (non-empty) hole by the bidirectional type
checker~\cite{popl-paper}.
%
Rather than trying to debug the error based on the static error, the programmer
may wish to trying running the function anyway by calling, say, \li{sum_list(2)}.
%
\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
%
By observing that the hole expression is being added to the integer \li{3}, he
realizes that it needs to be an integer, specifically, \li{0}.
%
Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
successfully proceeded to evaluate despite the error elsewhere.
\end{comment}

@@

\begin{comment}

@@

\end{comment}

### 2018-04-17 examples-3.tex
sec2!

\subsection{Live Programming with Gradual Type Errors}

\emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
programs that would otherwise be rejected by static type systems---either
because type inference cannot reconstruct a valid type assignment, or because
there may not be a unique valid type assignment at all.
%
In either case, gradual type systems allow types to contain \emph{unknown
types}, and partially unknown types can be used in contexts that require
different types, so long as they are \emph{consistent} (intuitively, equal
modulo the unknown parts).
%
Dynamic casts are then inserted to ensure that these remaining static
discrepancies are not violated at run-time.

@@

\overviewExample{5}{Dynamically Typed Negation}

@@

\noindent
%
As in prior gradually typed languages~\cite{XXX},
evaluating the expression \li{negate false 1} on line \rkc{XXX} leads to
\li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
lead to the \emph{cast error} \li{1<?? =/=> bool>} because \li{1} cannot be
safely treated as having type \li{bool}.
%
Unlike in prior gradually typed languages, however, \HazelnutLive{} evaluates
around the cast error, making progress on the \li{2 + 3 + negate false 4}
expression that surrounds the error; the final indeterminate result is
\li{(not (1<?? =/=> bool>))<bool => ??> + 1}.
%
Just like it is useful for static type checkers to report multiple errors, our
approach allows us to report multiple dynamic cast errors, and otherwise make
progress on expressions that do not depend on failed casts.
%
This approach can be incorporated into existing gradually typed languages.

>>>>

% !TEX root = hazelnut-dynamics.tex

\vspace{-4px}
\subsection{Example 4: Gradual Type Errors}
\vspace{-2px}
So far, we have only discussed programs where a hole appears within an expression.
\Hazel also allows holes to appear in types.
Fortunately, \citet{popl-paper} confirms that the substantial literature on \emph{gradual type systems} \cite{Siek06a,DBLP:conf/snapl/SiekVCB15} is directly relevant to the problem of reasoning statically around type holes.
Unsurprisingly, it is also relevant to the problem of running 
programs with incomplete types. Indeed, that is the very purpose of gradual typing. As such, let us consider only a small synthetic example to demonstrate what is different in \Hazel.
\begin{lstlisting}
f : ? -> int
f x = if x then 1 else 2
(f true, f 3, f false)
\end{lstlisting}
Evaluation produces the following result:
\begin{lstlisting}[numbers=none]
(1, if SSTR3<int => ? =/> bool>ESTR then 1 else 2, 2)
\end{lstlisting}
Here, we have left a hole, \li{?}, in the type of \li{f}. On Line 3, we call \li{f} three times, with two different types of arguments. The static semantics of gradual typing permits all of these calls (see next section) and indeed, the first call, with \li{true} as the argument, evaluates unproblematically to \li{1}.

The second call, however, is problematic: \li{3} is of type \li{int} but it flows into boolean guard position. The usual semantics for gradual typing would simply abort evaluation here with a dynamic type error. However, this deprives the programmer of live feedback about other parts of the program that might not depend at all on this problematic function call---in this case, the third call to \li{f}. Instead, we follow the approach we took above for static type inconsistencies: we treat a failed cast like a hole. A failed cast, notated textually and in red \li{SSTR3<int => ? =/> bool>ESTR} (see next section), reports both the expected type, here \li{bool}, and the assigned type, here \li{int}. Consequently, the second call to \li{f} stops as if the guard were an empty hole, but the third call can progress safely.

\begin{comment}
\emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
programs that would otherwise be rejected by static type systems---either
because type inference cannot reconstruct a valid type assignment, or because
there may not be a unique valid type assignment at all.
%
In either case, gradual type systems allow types to contain \emph{unknown
types}, and partially unknown types can be used in contexts that require
different types, so long as they are \emph{consistent} (intuitively, equal
modulo the unknown parts).
%
Dynamic casts are then inserted to ensure that these remaining static
discrepancies are not violated at run-time.

@@

\overviewExample{3}{Dynamically Typed Negation}

@@

\noindent
%
As in prior gradually typed languages~\cite{XXX},
evaluating the expression \li{negate false 1} on line \rkc{XXX} leads to
\li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
lead to the \emph{cast error} \li{1<?? =/=> bool>} because \li{1} cannot be
safely treated as having type \li{bool}.
%
Unlike in prior gradually typed languages, however, \HazelnutLive{} evaluates
around the cast error, making progress on the \li{2 + 3 + negate false 4}
expression that surrounds the error; the final indeterminate result is
\li{(not (1<?? =/=> bool>))<bool => ??> + 1}.
%
Just like it is useful for static type checkers to report multiple errors, our
approach allows us to report multiple dynamic cast errors, and otherwise make
progress on expressions that do not depend on failed casts.
%
This approach can be incorporated into existing gradually typed languages.
\end{comment}
